---
title: "Scoring model for Default of Credit Card Clients"
date: "`r Sys.Date()`"
author: Julita Cichońska
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE, out.width = 750)

```

# Intro

Our goal is to build a scoring model to assess probability of default of credit card clients. We are going to use a dataset which can be found here:

<https://archive.ics.uci.edu/ml/datasets/default+of+credit+card+clients>.


## Features Description

<details>
  <summary>Source reference</summary>
  
This research employed a binary variable, default payment (Yes = 1, No = 0), as the response variable. This study reviewed the literature and used the following 23 variables as explanatory variables:

X1: Amount of the given credit (NT dollar): it includes both the individual consumer credit and his/her family (supplementary) credit.

X2: Gender (1 = male; 2 = female).

X3: Education (1 = graduate school; 2 = university; 3 = high school; 4 = others).

X4: Marital status (1 = married; 2 = single; 3 = others).

X5: Age (year).

X6 - X11: History of past payment. We tracked the past monthly payment records (from April to September, 2005) as follows: X6 = the repayment status in September, 2005; X7 = the repayment status in August, 2005; . . .;X11 = the repayment status in April, 2005. The measurement scale for the repayment status is: -1 = pay duly; 1 = payment delay for one month; 2 = payment delay for two months; . . .; 8 = payment delay for eight months; 9 = payment delay for nine months and above.

X12-X17: Amount of bill statement (NT dollar). X12 = amount of bill statement in September, 2005; X13 = amount of bill statement in August, 2005; . . .; X17 = amount of bill statement in April, 2005.

X18-X23: Amount of previous payment (NT dollar). X18 = amount paid in September, 2005; X19 = amount paid in August, 2005; . . .;X23 = amount paid in April, 2005.
</details>


## Data Loading


```{r libraries}

library(readxl)
library(dplyr)
library(ggplot2)
library(purrr)
library(corrplot)
library(scorecard)
library(stringr)
library(kableExtra)
library(gridExtra)
source('utility_functions.R')

```

Firstly, we are going to load the dataset from xls file.

```{r read_data}

df_raw <- read_excel(path = 'default of credit card clients.xls', col_names = TRUE, skip = 1)

head(df_raw, 10) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = "hover") %>% 
  scroll_box(width = "800px", height = "400px")

```



Next, we need to change name of response variable and drop unnecesary column.


```{r}

colnames(df_raw)[25] <- 'DEFAULT'
df_raw$ID <- NULL

```


# Preprocessing


## Recoding


We will focus mainly on categorical features: SEX, EDUCATION and MARRIAGE. We will decode those features to original values, as we will transform variables with Weight of Evidence method later on.


```{r}

df <- df_raw %>% mutate(
                    SEX = as.factor(case_when(SEX == 1 ~ 'male',
                                              SEX == 2 ~ 'female')),
                    EDUCATION = as.factor(case_when(EDUCATION == 1 ~'graduate school',
                                                    EDUCATION == 2 ~'university',
                                                    EDUCATION == 3 ~'high school',
                                                    EDUCATION == 4 ~'others')),
                    MARRIAGE = as.factor(case_when(MARRIAGE == 1 ~ 'married',
                                                   MARRIAGE == 2 ~ 'single',
                                                   MARRIAGE == 3 ~ 'others')))

```


## Dataset review


```{r}

featuresReview(df) %>% kable() %>% kable_styling(bootstrap_options = "hover") %>% scroll_box(height = "400px")

```

## Missing Data Imputation

As we can see in the summary above, there are missing values in variables EDUCATION and MARRIAGE.

```{r}

imputeFactor <- function(v_col, string_imp = 'missing'){
  levels(v_col) <- c(levels(v_col) , string_imp)
  v_col[is.na(v_col)] <- string_imp
  v_col
}

df$EDUCATION <- imputeFactor(df$EDUCATION)
df$MARRIAGE <- imputeFactor(df$MARRIAGE)

```

## Variables Types

```{r}

df <- df %>% mutate(across(starts_with("PAY")&!contains("AMT"), as.integer))
df <- df %>% mutate(DEFAULT = as.integer(DEFAULT))

```

## Learn Test Split

```{r}
set.seed(2021)
n_obs <- dim(df)[1]
idx <- sample(1:n_obs, size = 0.75*n_obs, replace = FALSE)

df_learn <- df[idx,]
df_test <- df[-idx,]

```

Default Rate after split

```{r}
data.frame(set = c('learn', 'test'),
                   default_rate = rbind(mean(df_learn$DEFAULT),mean(df_test$DEFAULT))
) %>% kable() %>% kable_styling(full_width = F)
```


# Features Analysis

## Distributions


Selecting names of variables to plot

```{r asis = TRUE}

df_learn_fct <- df_learn %>% select_if(funs(is.factor(.)||is.integer(.)))
df_learn_cont <- df_learn %>% select_if(funs(!is.factor(.)&!is.integer(.)))
```



```{r}
plots_list <- plotFactor(df_learn_fct, colnames(df_learn_fct)) 
for (i in seq_along(plots_list)){
  plots_list[[i]] <- list(plots_list[[i]], colnames(df_learn_fct)[i])}
```

## Categorical and discrete variables {.tabset}

```{r results = 'asis'}
for (i in seq_along(plots_list)){
  tmp <- plots_list[[i]]
  cat('###', tmp[[2]], " \n")
  print(tmp[[1]])
  cat('\n\n')}
```


```{r}
plots_list <- plotHistogram(df_learn_cont, colnames(df_learn_cont) )
for (i in seq_along(plots_list)){
  plots_list[[i]] <- list(plots_list[[i]], colnames(df_learn_cont)[i])}
```

## Continuous variables {.tabset}

```{r results = 'asis'}
for (i in seq_along(plots_list)){
  tmp <- plots_list[[i]]
  cat('###', tmp[[2]], " \n")
  print(tmp[[1]])
  cat('\n\n')}
```

## Correlations

Selection of numerical variables, correlation matrix, corrplot with clusters

```{r}
df_learn_num <- df_learn %>% select_if(is.numeric) %>% select(-DEFAULT)

cor_M <- cor(df_learn_num)

cor_plot <- corrplot(corr = cor_M,
                     method = 'circle', 
                     type = 'full', 
                     order = 'hclust', 
                     hclust.method = 'complete',
                     addrect = 12, 
                     rect.col = 'tomato3',
                     tl.col = 'grey20', 
                     tl.cex = 0.6, 
                     tl.srt = 45)
```


Clusters extraction: hclust on 1-cor_M and cutting a dendrogram


```{r}

clust_obj <- hclust(as.dist(1-cor_M))

plot(clust_obj)

clust_cut <- data.frame(cluster = cutree(clust_obj, k = 12)) %>% mutate(variable = rownames(.))

```


Selection of one representative from every cluster based on information value

```{r}

df_iv <- data.frame(scorecard::iv(df_learn %>% select_if(is.numeric), 'DEFAULT'))
 
df_merge <- merge(clust_cut, df_iv, by = 'variable')

df_merge %>% 
  arrange(cluster, -info_value) %>% 
  kable() %>% kable_styling(full_width = F, bootstrap_options = "hover") %>% scroll_box(height = '300px', width = '300px')

```


```{r}
df_sel_numeric <- df_merge %>% 
  group_by(cluster) %>% 
  arrange(-info_value) %>% 
  top_n(1) %>% arrange(cluster)

names_sel_num <- df_sel_numeric %>% pull(variable)

```

# Weight of Evidence (WoE) transformation

Preparing data frame for Weight of Evidence tranformation

```{r}

names_sel_all <- c(names_sel_num, names(df_learn_fct),  'DEFAULT')

df_woe <- df_learn %>% select(all_of(names_sel_all))

df_woebin <- scorecard::woebin(df_woe, 'DEFAULT')

df_learn_woe <- scorecard::woebin_ply(df_woe, bins = df_woebin )

```


```{r}
plots_list <- plotWoe(df_woebin, names(df_woebin) )
for (i in seq_along(plots_list)){
  plots_list[[i]] <- list(plots_list[[i]], names(df_woebin)[i])}
```

## WoEPlots {.tabset}

```{r results = 'asis'}
for (i in seq_along(plots_list)){
  tmp <- plots_list[[i]]
  cat('###', tmp[[2]], " \n")
  print(tmp[[1]])
  cat('\n\n')}
```


# Model

## Correlations check

```{r}

cor_M2 <- cor(df_learn_woe )
cor_plot2 <- corrplot(cor_M2,
                     method = 'circle', 
                     type = 'full', 
                     order = 'hclust', 
                     tl.col = 'grey20', 
                     tl.cex = 0.6, 
                     tl.srt = 45)

```

## Balancing dataset

```{r}

weights_vec <- ifelse(df_learn_woe$DEFAULT == 1, 4, 1)

```


## Building Logistic Regression Model

```{r}
vars_names <- c(      
"LIMIT_BAL_woe"
#,"AGE_woe"       
,"PAY_0_woe"     
#,"PAY_2_woe"     
,"PAY_4_woe"    
#,"BILL_AMT6_woe" 
,"PAY_AMT1_woe"  
,"PAY_AMT2_woe"  
#,"PAY_AMT3_woe"  
#,"PAY_AMT4_woe"  
,"PAY_AMT5_woe" 
,"PAY_AMT6_woe"  
,"SEX_woe"       
#,"EDUCATION_woe" 
,"MARRIAGE_woe"  
)

formula <- as.formula(paste("DEFAULT ~ ", paste(vars_names, collapse= "+")))

model <- glm(df_learn_woe, 
             formula = formula, 
             family = 'binomial', 
             weights =  weights_vec)

summary(model)

```

## VIF

```{r}
vif(model)
```


## Prediction

Transforming test set and gathering results in one data frame with original variables, WoE transformed variables and prediction.

```{r result = 'hide'}

# LEARN_SET
df_learn_final <- cbind(df_learn  %>% select(-DEFAULT), df_learn_woe)
df_learn_final$y_pred <- model$fitted.values
df_learn_final$set <- 'learn'

# TEST_SET
df_test_sel <- df_test %>% select(all_of(c(names(df_woebin), 'DEFAULT')))
df_test_woe <- scorecard::woebin_ply(df_test_sel, bins = df_woebin )
df_test_final <- cbind(df_test %>% select(-DEFAULT), df_test_woe)
df_test_final$y_pred <- predict(model, df_test_woe, type = 'response' )
df_test_final$set <- 'test'

# FINAL_SET
df_final <- rbind(df_learn_final, df_test_final)
df_final$y_pred_bin <- cut(
  df_final$y_pred, 
  breaks = quantile(df_final$y_pred, probs = seq(0, 1, by = 0.1)), 
  include.lowest = TRUE, 
  right = TRUE)

```



## Diagnostics

```{r}
df_final %>% 
  group_by(set) %>% 
  summarise(
    Gini = MLmetrics::Gini(y_pred, DEFAULT),
    AUC = MLmetrics::AUC(y_pred, DEFAULT)
  ) %>% 
  kable() %>% kable_styling(full_width = F)

plotGoodBadDist(df_final, y_pred = 'y_pred', y_real = 'DEFAULT', set_var = 'set' )
plotSuccessRateInBins(df_final,'y_pred_bin','DEFAULT', 'set', 'Default Rate' )
plotROC(df_final$DEFAULT,df_final$y_pred, n_cutoff = 100, df_final$set)
plotGainChart(df_final, 'y_pred', 'DEFAULT', 'set')


```

```{r}
df_list <- scorecard(df_woebin, model)

for (i in seq_along(df_list)){
  df_list[[i]] <- list(df_list[[i]], names(df_list)[i])}
```

## Scorecard {.tabset}

```{r results = 'asis'}
for (i in seq_along(df_list)){
  tmp <- df_list[[i]]
  cat('###', tmp[[2]], " \n")
  print(data.frame(tmp[[1]]) %>%kable() %>% kable_styling() %>% scroll_box(width = '750px'))
  cat('\n\n')}
```



# Random Forest approach

```{r}
library(ranger)

ranger_model <- ranger(DEFAULT ~., 
                       data = df_learn, 
                       probability = TRUE, 
                       max.depth = 7, 
                       case.weights = weights_vec)

df_ranger_learn <- df_learn
df_ranger_learn$y_pred <- ranger_model$predictions[,2]
df_ranger_learn$set <- 'learn'


df_ranger_test <- df_test
df_ranger_test$y_pred <- predict(ranger_model, data = df_test, type = "response")$prediction[,2]
df_ranger_test$set <- 'test'

df_ranger_final <- rbind(df_ranger_learn, df_ranger_test)

df_ranger_final$y_pred_bin <- cut(df_ranger_final$y_pred, 
                                  breaks = quantile(df_ranger_final$y_pred,
                                                    probs = seq(0, 1, by = 0.1)), 
                                  include.lowest = TRUE, 
                                  right = TRUE)

```

## Diagnostics 


```{r}
df_ranger_final %>% 
  group_by(set) %>% 
  summarise(
    Gini = MLmetrics::Gini(y_pred, DEFAULT),
    AUC = MLmetrics::AUC(y_pred, DEFAULT)
  ) %>% 
  kable() %>% kable_styling(full_width = F)

plotGoodBadDist(df_ranger_final,'y_pred', 'DEFAULT', 'set' )
plotSuccessRateInBins(df_ranger_final,'y_pred_bin','DEFAULT', 'set', 'Default Rate' )
plotROC(df_ranger_final$DEFAULT,df_ranger_final$y_pred, n_cutoff = 100, df_ranger_final$set)
plotGainChart(df_ranger_final, 'y_pred', 'DEFAULT', 'set')

```





