---
title: "Scoring model for Default of Credit Card Clients"
date: "`r Sys.Date()`"
author: Julita Cichońska
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE, out.width = 750)

```

# Intro

Our goal is to build a scoring model to assess probability of default of credit card clients. We are going to use a dataset which can be found here:

<https://archive.ics.uci.edu/ml/datasets/default+of+credit+card+clients>.


## Features Description

<details>
  <summary>Source reference</summary>
  
This research employed a binary variable, default payment (Yes = 1, No = 0), as the response variable. This study reviewed the literature and used the following 23 variables as explanatory variables:

X1: Amount of the given credit (NT dollar): it includes both the individual consumer credit and his/her family (supplementary) credit.

X2: Gender (1 = male; 2 = female).

X3: Education (1 = graduate school; 2 = university; 3 = high school; 4 = others).

X4: Marital status (1 = married; 2 = single; 3 = others).

X5: Age (year).

X6 - X11: History of past payment. We tracked the past monthly payment records (from April to September, 2005) as follows: X6 = the repayment status in September, 2005; X7 = the repayment status in August, 2005; . . .;X11 = the repayment status in April, 2005. The measurement scale for the repayment status is: -1 = pay duly; 1 = payment delay for one month; 2 = payment delay for two months; . . .; 8 = payment delay for eight months; 9 = payment delay for nine months and above.

X12-X17: Amount of bill statement (NT dollar). X12 = amount of bill statement in September, 2005; X13 = amount of bill statement in August, 2005; . . .; X17 = amount of bill statement in April, 2005.

X18-X23: Amount of previous payment (NT dollar). X18 = amount paid in September, 2005; X19 = amount paid in August, 2005; . . .;X23 = amount paid in April, 2005.
</details>


## Data Loading


```{r libraries}

library(readxl)
library(dplyr)
library(ggplot2)
library(purrr)
library(corrplot)
library(scorecard)
library(stringr)
library(kableExtra)
library(gridExtra)
source('utility_functions.R')

```

Firstly, we are going to load the dataset from xls file.

```{r read_data}

df_raw <- read_excel(path = 'default of credit card clients.xls', col_names = TRUE, skip = 1)

head(df_raw, 10) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = "hover") %>% 
  scroll_box(width = "800px", height = "400px")

```



Next we need to change name of response variable and drop unnecesary column.


```{r}

colnames(df_raw)[25] <- 'DEFAULT'
df_raw$ID <- NULL

```


# Preprocessing


## Recoding


We will focus mainly on categorical features: SEX, EDUCATION and MARRIAGE. We will decode those features to original values, as we will transform variables with Weight of Evidence method later on.


```{r}

df <- df_raw %>% mutate(
                    SEX = as.factor(case_when(SEX == 1 ~ 'male',
                                              SEX == 2 ~ 'female')),
                    EDUCATION = as.factor(case_when(EDUCATION == 1 ~'graduate school',
                                                    EDUCATION == 2 ~'university',
                                                    EDUCATION == 3 ~'high school',
                                                    EDUCATION == 4 ~'others')),
                    MARRIAGE = as.factor(case_when(MARRIAGE == 1 ~ 'married',
                                                   MARRIAGE == 2 ~ 'single',
                                                   MARRIAGE == 3 ~ 'others')))

```


## Dataset review


```{r}

featuresReview(df) %>% kable() %>%
  kable_styling(bootstrap_options = "hover") %>% 
  scroll_box(height = "400px")

```

## Missing Data Imputation

As we can see in the summary above, there are missing values in variables EDUCATION and MARRIAGE.

```{r}

imputeFactor <- function(v_col, string_imp = 'missing'){
  levels(v_col) <- c(levels(v_col) , string_imp)
  v_col[is.na(v_col)] <- string_imp
  v_col
}

df$EDUCATION <- imputeFactor(df$EDUCATION)
df$MARRIAGE <- imputeFactor(df$MARRIAGE)

```

## Variables Types

Finally, we need to cast some variables to integers.

```{r}

df <- df %>% mutate(across(starts_with("PAY")&!contains("AMT"), as.integer))
df <- df %>% mutate(DEFAULT = as.integer(DEFAULT))

```

## Learn Test Split

Let's split data set into two parts: learn set and test set in proportion 3:1.

```{r}
set.seed(2021)
n_obs <- dim(df)[1]
idx <- sample(1:n_obs, size = 0.75*n_obs, replace = FALSE)

df_learn <- df[idx,]
df_test <- df[-idx,]

```

Checking Default Rate after split - need to be roughly equal.

```{r}
data.frame(set = c('learn', 'test'),
                   default_rate = rbind(mean(df_learn$DEFAULT),mean(df_test$DEFAULT))
) %>% kable() %>% kable_styling(full_width = F)
```


# Features Analysis

## Distributions


Selecting names of variables to plot

```{r asis = TRUE}

df_learn_fct <- df_learn %>% select_if(funs(is.factor(.)||is.integer(.)))
df_learn_cont <- df_learn %>% select_if(funs(!is.factor(.)&!is.integer(.)))
```




## Categorical and discrete variables {.tabset}

Plotting distributions of categorical and descrete variables using utility function plotFactor (utility_functions.R)

```{r}
plots_list <- plotFactor(df_learn_fct, colnames(df_learn_fct)) 
for (i in seq_along(plots_list)){
  plots_list[[i]] <- list(plots_list[[i]], colnames(df_learn_fct)[i])}
```

```{r results = 'asis'}
for (i in seq_along(plots_list)){
  tmp <- plots_list[[i]]
  cat('###', tmp[[2]], " \n")
  print(tmp[[1]])
  cat('\n\n')}
```



## Continuous variables {.tabset}

Plotting distributions of continuous variables using utility function plotHistogram (utility_functions.R)

```{r}
plots_list <- plotHistogram(df_learn_cont, colnames(df_learn_cont) )
for (i in seq_along(plots_list)){
  plots_list[[i]] <- list(plots_list[[i]], colnames(df_learn_cont)[i])}
```


```{r results = 'asis'}
for (i in seq_along(plots_list)){
  tmp <- plots_list[[i]]
  cat('###', tmp[[2]], " \n")
  print(tmp[[1]])
  cat('\n\n')}
```

## Correlations

In order to examine possible correlations between variables we are selecting numerical variables and calculating correlation matrix. To visualize correlation matrix we are using corrplot with hierachical clustering to recognize groups of most correlated variables.


```{r}
df_learn_num <- df_learn %>% select_if(is.numeric) %>% select(-DEFAULT)

cor_M <- cor(df_learn_num)

cor_plot <- corrplot(corr = cor_M,
                     method = 'circle', 
                     type = 'full', 
                     order = 'hclust', 
                     hclust.method = 'complete',
                     addrect = 12, 
                     rect.col = 'tomato3',
                     tl.col = 'grey20', 
                     tl.cex = 0.6, 
                     tl.srt = 45)
```



Clusters extraction: plotting cluster dendrogram with parameters matching hclust in corrplot (distance of 1-cor_M) and cutting dendrogram to specific number of clusters.

```{r}

clust_obj <- hclust(as.dist(1-cor_M))

plot(clust_obj)

clust_cut <- data.frame(cluster = cutree(clust_obj, k = 12)) %>% mutate(variable = rownames(.))

```

## Feature selection

Finally, we are selecting one representative from every cluster based on information value.

```{r}

df_iv <- data.frame(scorecard::iv(df_learn %>% select_if(is.numeric), 'DEFAULT'))
 
df_merge <- merge(clust_cut, df_iv, by = 'variable')

df_merge %>% 
  arrange(cluster, -info_value) %>% 
  kable() %>% 
  kable_styling(full_width = F, bootstrap_options = "hover") %>% scroll_box(height = '300px', width = '300px')

```


```{r}
df_sel_numeric <- df_merge %>% 
  group_by(cluster) %>% 
  arrange(-info_value) %>% 
  top_n(1) %>% arrange(cluster)

names_sel_num <- df_sel_numeric %>% pull(variable)

```

# WoE transformation

Preparing data frame for Weight of Evidence tranformation

```{r}

names_sel_all <- c(names_sel_num, names(df_learn_fct),  'DEFAULT')

df_woe <- df_learn %>% select(all_of(names_sel_all))

df_woebin <- scorecard::woebin(df_woe, 'DEFAULT')

df_learn_woe <- scorecard::woebin_ply(df_woe, bins = df_woebin )

```

## WoEPlots {.tabset}

```{r}
plots_list <- plotWoe(df_woebin, names(df_woebin) )
for (i in seq_along(plots_list)){
  plots_list[[i]] <- list(plots_list[[i]], names(df_woebin)[i])}
```



```{r results = 'asis'}
for (i in seq_along(plots_list)){
  tmp <- plots_list[[i]]
  cat('###', tmp[[2]], " \n")
  print(tmp[[1]])
  cat('\n\n')}
```


# Model

## Correlations check

```{r}

cor_M2 <- cor(df_learn_woe )
cor_plot2 <- corrplot(cor_M2,
                     method = 'circle', 
                     type = 'full', 
                     order = 'hclust', 
                     tl.col = 'grey20', 
                     tl.cex = 0.6, 
                     tl.srt = 45)

```

We can see, that there are still quite correlated variables, but for now this is sufficient. Later we will examine Variance Inflation Factor after building the model.

## Balancing dataset

Due to unequal proportion of Default and Non-Default we need to balance dataset by adding weights to every observation.

```{r}

weights_vec <- ifelse(df_learn_woe$DEFAULT == 1, 4, 1)

```


## Building Logistic Regression Model

Due to small number of variables at this point, there is no need to use step function to perform feature selection. We will use error and trial method to select most informative variables. The most importat part is to be careful with estimated coefficient: because variables are transformed by WoE, coefficients should be all positive. Negative coefficient would suggest reverse relationship between predictor and response variable from what was obtained during WoE transformation.

```{r}
vars_names <- c(      
"LIMIT_BAL_woe"
#,"AGE_woe"       
,"PAY_0_woe"     
#,"PAY_2_woe"     
,"PAY_4_woe"    
#,"BILL_AMT6_woe" 
,"PAY_AMT1_woe"  
,"PAY_AMT2_woe"  
#,"PAY_AMT3_woe"  
#,"PAY_AMT4_woe"  
,"PAY_AMT5_woe" 
,"PAY_AMT6_woe"  
,"SEX_woe"       
#,"EDUCATION_woe" 
,"MARRIAGE_woe"  
)

formula <- as.formula(paste("DEFAULT ~ ", paste(vars_names, collapse= "+")))

model <- glm(df_learn_woe, 
             formula = formula, 
             family = 'binomial', 
             weights =  weights_vec)

summary(model)

```

## VIF

Finally, last check for some collinearity of variables by calculating Variance Inflation Factor.

```{r}
vif(model)
```

Vifs below 1.5 suggests very small amount of collinearity (anything above 2 should be suspicious).

## Prediction

Next, we will transform test set and gather results in one data frame with original variables, WoE transformed variables and prediction.

```{r result = 'hide'}

# LEARN_SET
df_learn_final <- cbind(df_learn  %>% select(-DEFAULT), df_learn_woe)
df_learn_final$y_pred <- model$fitted.values
df_learn_final$set <- 'learn'

# TEST_SET
df_test_sel <- df_test %>% select(all_of(c(names(df_woebin), 'DEFAULT')))
df_test_woe <- scorecard::woebin_ply(df_test_sel, bins = df_woebin )
df_test_final <- cbind(df_test %>% select(-DEFAULT), df_test_woe)
df_test_final$y_pred <- predict(model, df_test_woe, type = 'response' )
df_test_final$set <- 'test'

# FINAL_SET
df_final <- rbind(df_learn_final, df_test_final)
df_final$y_pred_bin <- cut(
  df_final$y_pred, 
  breaks = quantile(df_final$y_pred, probs = seq(0, 1, by = 0.1)), 
  include.lowest = TRUE, 
  right = TRUE)


head(df_final) %>% kable() %>% 
  kable_styling(bootstrap_options = "hover") %>%
  scroll_box(width = "800px", height = "400px")

```


## Diagnostics

For diagnostics we will use utility functions which I've created to show distributions of Goods/Bads, ROC curve and Gain Chart.

```{r}
df_final %>% 
  group_by(set) %>% 
  summarise(
    Gini = MLmetrics::Gini(y_pred, DEFAULT),
    AUC = MLmetrics::AUC(y_pred, DEFAULT)
  ) %>% 
  kable() %>% kable_styling(full_width = F)

plotGoodBadDist(df_final, y_pred = 'y_pred', y_real = 'DEFAULT', set_var = 'set' )
plotSuccessRateInBins(df_final,'y_pred_bin','DEFAULT', 'set', 'Default Rate' )
plotROC(df_final$DEFAULT,df_final$y_pred, n_cutoff = 100, df_final$set)
plotGainChart(df_final, 'y_pred', 'DEFAULT', 'set')


```

```{r}
df_list <- scorecard(df_woebin, model)

for (i in seq_along(df_list)){
  df_list[[i]] <- list(df_list[[i]], names(df_list)[i])}
```

The result is quite satisfying. Model didn't overfit, has good predictive power and is easy to explain, which is very important concerning nature of business problem (risk among credit cards clients). Model can help in recognizing potential defaulting clients, but there is a large group of Bad Clients (Default = 1), which haven't been spotted (received low probability of default: peak around probability of 0.4), which is a space for further improvement. To summarise we will create scorecard for this model.



## Scorecard {.tabset}

```{r results = 'asis'}
for (i in seq_along(df_list)){
  tmp <- df_list[[i]]
  cat('###', tmp[[2]], " \n")
  print(data.frame(tmp[[1]]) %>%kable() %>% kable_styling() %>% scroll_box(width = '750px'))
  cat('\n\n')}
```



# BONUS: Random Forest approach

To establish reference point for our logistic regression model, we will quickly build random forest model (using very fast ranger package) for the same data (before WoE transformation) with some arbitrarily chosen hyperparameters, with no hyperparameter tuning.

```{r}
library(ranger)

ranger_model <- ranger(DEFAULT ~., 
                       data = df_learn, 
                       probability = TRUE, 
                       max.depth = 7, 
                       min.node.size = 20,
                       mtry= 5,
                       case.weights = weights_vec)

df_ranger_learn <- df_learn
df_ranger_learn$y_pred <- ranger_model$predictions[,2]
df_ranger_learn$set <- 'learn'


df_ranger_test <- df_test
df_ranger_test$y_pred <- predict(ranger_model, data = df_test, type = "response")$prediction[,2]
df_ranger_test$set <- 'test'

df_ranger_final <- rbind(df_ranger_learn, df_ranger_test)

df_ranger_final$y_pred_bin <- cut(df_ranger_final$y_pred, 
                                  breaks = quantile(df_ranger_final$y_pred,
                                                    probs = seq(0, 1, by = 0.1)), 
                                  include.lowest = TRUE, 
                                  right = TRUE)

```

## Diagnostics 


```{r}
df_ranger_final %>% 
  group_by(set) %>% 
  summarise(
    Gini = MLmetrics::Gini(y_pred, DEFAULT),
    AUC = MLmetrics::AUC(y_pred, DEFAULT)
  ) %>% 
  kable() %>% kable_styling(full_width = F)

plotGoodBadDist(df_ranger_final,'y_pred', 'DEFAULT', 'set' )
plotSuccessRateInBins(df_ranger_final,'y_pred_bin','DEFAULT', 'set', 'Default Rate' )
plotROC(df_ranger_final$DEFAULT,df_ranger_final$y_pred, n_cutoff = 100, df_ranger_final$set)
plotGainChart(df_ranger_final, 'y_pred', 'DEFAULT', 'set')

```

Model resulted in Gini close to 0.54 on test set, which is significantly better comparing to 0.51 for previous model. Random forest model is slightly overfitted, so there is potential for improvement, so the result for test set could be even better. There is enhancement in recognizing Bad Clients: peak of not recognized Bad Clients is slightly lower than before and it could suggest, that ensembled tree method can recognize some pattern which wasn't detectable in linear model built on WoE transformed data (which itself is huge loss of information).



